# Time Series Analysis of Air Pollution and Health Data

## Initialization

Loading required packages. 
```{r}
if(!require(tsModel)) {
        install.packages("tsModel")
}
if(!require(moments)) {
        install.packages("moments")
}
library(splines)
library(methods)
source("../R/datastructures.R")
```

## Read in Data

Read in the raw data.

```{r}
d0 <- readAPTSData("ny.rds")
```

## Exploratory Data Analysis

Run the exploratory analysis.

```{r}
ch <- checkData(d0)
print(ch)
```


## Model Building

Fitting the base model.
```{r}
base <- glm(death ~ dow + ns(date, 19 * 6) + ns(tmpd, 6) + ns(rmtmpd, 6) + ns(dptp, 3) + ns(rmdptp, 3) + Lag(pm10tmean, 1), data = as(d0, "data.frame"), family = poisson)
```

Selecting the optimal degrees of freedom for the smooth function of
time to adjust for unmeasured confounding.

```{r}
ap <- new("APTSModel", timevar = "date", model = base, exposure = "pm10tmean", summary = ch)
dfmin <- selectTimeDF(ap)
```

Optimal number of degrees of freedom _per year_ for the smooth
function of time.

```{r}
show(dfmin)
```

Check AIC values for selecting the degrees of freedom per year.

```{r}
plot(dfmin)
```

## Model Fitting

Fitting the final model.
```{r}
df <- as(dfmin, "numeric")
final <- glm(death ~ dow + ns(date, 19 * df) + ns(tmpd, 6) + ns(rmtmpd, 6) + ns(dptp, 3) + ns(rmdptp, 3) + Lag(pm10tmean, 1), data = as(d0, "data.frame"), family = poisson)
ap <- new("APTSModel", timevar = "date", model = final, exposure = "pm10tmean", summary = ch)
```

## Results

Final results for a 10 unit change in exposure.

```{r}
writeLines(strwrap(modelEffect(ap, incr = 10, pvalue = TRUE)))
```

## Sensitivity Analysis

Run sensitivity analysis with respect to the degrees of freedom for
the smooth function of time.

```{r}
sens <- sensitivityAnalysis(ap, 4:20)
show(sens)
plot(sens)
```

Check sensitivity of results to outliers or influential data points.

```{r}
infl <- influenceAnalysis(ap)
plot(infl)
```


## Appendix

Summary of final fitted model.

```{r}
summary(final)
```
